---
title: "Sleep_Confidence"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Are Students Getting Enough Sleep?

The CDC recommends that adults aged 18-25 get at least 7 hours of sleep per night. A campus health reporter wants to know: are students at this university getting enough sleep?

Remember providing the hours you slept the night before our first class? Your job is to analyze this data and determine what we can confidently say about sleep habits across the entire student body.

This is the core challenge of confidence intervals: using a **sample** to make claims about a **population**.

### Task 1: Load the tidyverse and the data

```{r}
library(tidyverse)

sleep_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/sleep_data.csv")

head(sleep_data)
```

### Task 2: Calculate descriptive statistics

Let's start with what we know about these students. Calculate the mean, standard deviation and number of observations.

```{r}
sleep_stats <- sleep_data |>
  summarise(
    mean_hours = mean(hours),
    sd_hours = sd(hours),
    n = n()
  )

sleep_stats
```

**Answer this question:** Based on these numbers alone, what would you tell the health reporter? Why might it be risky to report the average as the definitive answer for all students?

The sample size on this group is particularly small and overrepresents students from specific campus communities, such as the school of journalism. Communities like graduate students, underclassmen, non-journalism majors etc. are deeply underrepresented here and would potentially have vastly different results to report. 

### Task 3: Calculate the standard error

The standard error tells us how much our sample mean would vary if we took many different samples of 37 students. It's the key to moving from "this is what 37 students told us" to "this is what we think is true for the whole campus." The formula is: standard deviation divided by the square root of the sample size.

```{r}
sleep_stats <- sleep_stats |>
  mutate(
    se_hours = sd_hours / sqrt(37)
  )

sleep_stats
```

**Answer this question:** The standard error is much smaller than the standard deviation. Why? What does each one measure?

The standard deviation represents the total spread of the data by measuring how much results vary from the mean. Standard error is ONLY taking into account how much the mean itself would vary, not taking into account the spread of the overall data. The standard deviation is considering more total data, which would cause it to have a wider range. 

### Task 4: Calculate a 95% confidence interval

Now we'll build our confidence interval. For a 95% CI with a small sample, we use the t-distribution. The `qt(0.975, df = n - 1)` function gives us the critical value — the number of standard errors we go above and below the mean. We multiply critical value by the standard error to get the margin of error, then add/subtract from the mean.

```{r}
sleep_stats <- sleep_stats |>
  mutate(
    t_critical = qt(0.975, df = n - 1),
    margin_error = t_critical * se_hours,
    ci_lower = mean_hours - margin_error,
    ci_upper = mean_hours + margin_error
  )

sleep_stats |>
  select(mean_hours, margin_error, ci_lower, ci_upper)
```

**Answer this question:** Write one sentence explaining what this confidence interval means in plain English that a reader would understand.

The confidence interval shows the range in which the true mean more than likely falls based on what we know about the margin of error and the standard error. In this case, it shows that there is a 95% chance that the mean of hours falls between 7.5 and 7.9. The margin of error shows that these results could be incorrect by up to 4.5% in either direction.  

### Task 5: Visualize the distribution with the confidence interval

This chart shows the histogram of individual sleep hours, plus the mean (red line) and the 95% confidence interval (blue shaded region). The shaded region is the range where we think the true average for all students falls.

```{r}
ggplot(sleep_data, aes(x = hours)) +
  geom_histogram(binwidth = 0.5, fill = "skyblue", color = "white", alpha = 0.7) +
  geom_vline(xintercept = sleep_stats$mean_hours, color = "red", linewidth = 1) +
  geom_vline(xintercept = sleep_stats$ci_lower, color = "blue", linetype = "dashed") +
  geom_vline(xintercept = sleep_stats$ci_upper, color = "blue", linetype = "dashed") +
  annotate("rect",
           xmin = sleep_stats$ci_lower,
           xmax = sleep_stats$ci_upper,
           ymin = 0, ymax = Inf,
           fill = "blue", alpha = 0.1) +
  annotate("text", x = sleep_stats$mean_hours + 0.1, y = 8,
           label = paste0("Mean: ", round(sleep_stats$mean_hours, 2)),
           hjust = 0, color = "red", fontface = "bold") +
  labs(
    title = "Distribution of Student Sleep Hours",
    subtitle = paste0("95% Confidence Interval for the mean: [",
                     round(sleep_stats$ci_lower, 2), ", ",
                     round(sleep_stats$ci_upper, 2), "]"),
    x = "Hours of Sleep",
    y = "Number of Students"
  ) +
  theme_minimal()
```

**Answer this question:** Look at the chart. Notice that the confidence interval is much narrower than the spread of individual data points. Why? What's the difference between the variation in individual sleep hours and the uncertainty about the average?

Because the sample sizes we are taking for this are so small, it is unlikely that the mean will ever fall outside of that point despite the large range of data points represented. The distribution of this sample size is fairly standard; thus, most of the data sets from identical sample sizes are likely to look similar and therefore have a similar mean. 

### Task 6: The CDC question — is 7 hours within the interval?

The CDC recommends at least 7 hours. Based on the confidence interval, can we say students are getting enough sleep? Let's check.

```{r}
cdc_recommendation <- 7

sleep_stats |>
  mutate(
    meets_cdc = if_else(ci_lower >= cdc_recommendation, "Yes - entire CI is at or above 7", "No - CI includes values below 7")
  ) |>
  select(mean_hours, ci_lower, ci_upper, meets_cdc)
```

**Answer this question:** Based on this result, which of these headlines is justified? Why?

  a) "Students Average More Than 7 Hours of Sleep"
  b) "Students May Not Be Getting Enough Sleep, Survey Suggests"
  c) "Survey: Students Getting Recommended Amount of Sleep"

A would be the most accurate considering that the entire confidence interval falls above 7, so students are averaging more than 7 hours of sleep no matter what based on our data. This does NOT mean that all students are getting more than 7 hours of sleep, because we know about the standard deviation of data points from this set. Thus, saying that "students are getting the recommended amount of sleep" is too much of a generalization about students as a whole. 

### Task 7: What about a bigger sample?

The margin of error depends on the sample size. What if we had surveyed 100 students instead of 37 and got the same mean and standard deviation? Let's recalculate with a hypothetical sample size 100.

```{r}
hypothetical_n <- 100

hypothetical_stats <- tibble(
  mean_hours = sleep_stats$mean_hours,
  sd_hours = sleep_stats$sd_hours,
  n = hypothetical_n,
  se_hours = sd_hours / sqrt(hypothetical_n),
  t_critical = qt(0.975, df = hypothetical_n - 1),
  margin_error = t_critical * se_hours,
  ci_lower = mean_hours - margin_error,
  ci_upper = mean_hours + margin_error
)

# Compare the two intervals side by side
comparison <- tibble(
  sample_size = c(sleep_stats$n, hypothetical_n),
  margin_error = c(sleep_stats$margin_error, hypothetical_stats$margin_error),
  ci_lower = c(sleep_stats$ci_lower, hypothetical_stats$ci_lower),
  ci_upper = c(sleep_stats$ci_upper, hypothetical_stats$ci_upper)
)

comparison
```

**Answer this question:** How did the margin of error change? What does this tell you about the relationship between sample size and certainty?

The margin of error in this case is much smaller, which tells us that we can calculate averages with a higher level of certainty when we have a larger sample size to go off of. 

### Task 8: Journalistic interpretation

You now have all the statistics you need. Based on what you've calculated:

Write a headline and a one-paragraph lede for a campus news story about student sleep habits. Your text should accurately reflect the confidence interval — don't overstate or understate the findings.

Average student sleeps over 7 hours, study finds 

A class study from the University of Maryland found that students in the Philip Merrill College of Journalism tend to sleep more than 7 hours on average - a statistic that might be shocking to those who remember long nights at the student paper. 
