---
title: "UMD Fees"
---

Follow along with the demo to learn how to calculate percentage change using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.


### Task 1: Load the tidyverse


```{r}
library(tidyverse)
```


### Task 2: Load the UMD Fees dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv

```{r}
umd_fees <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv")
```

### Task 3: Calculate the percentage change in each category between Fall 2024 and Fall 2021 using `mutate()`

```{r}
umd_fees_pct <- umd_fees |> mutate(pct_change = (`Fall 2024`- `Fall 2021`)/`Fall 2021` * 100)
```



```{r}
umd_fees_pct <- umd_fees |> mutate(pct_change = (`Fall 2025`- `Fall 2024`)/`Fall 2024` * 100)

```



### Task 4: Describe our results

Write a few sentences about what you found in the data. What's the story here?

Student sustainability is the fee with the most dramatic percentage raise from 2021 to 2025, but the raise is only a total of 18 dollars; the most significant raise in terms of dollar amounts is the increase in shuttle bus fees of over $100. When comparing 2024 to 2025, however, there is a significant spike in student activities fees that could be worth looking into. I also find it interesting that the student athletics fee has not changed at all, as well as the recently added fee of the Student Counseling Center, and would find consider of those good leads for potential stories. 


When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
