---
title: "WMATA_Ridership"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Quick Review: Making a Line Chart

Let's do a quick refresher on line charts using WMATA ridership data, since a line chart will be part of your homework. This is daily bus and rail data from January 2024 through February 2025 (you may recall it from the midterm). We'll aggregate it to weekly totals so the chart is easier to read, then make a line chart.

### Task 1: Load the tidyverse and the data

```{r}
library(tidyverse)
library(lubridate)

wmata <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")

head(wmata)
```

### Task 2: Make weekly totals

Daily data is noisy — weekends and holidays create big dips that make trends hard to see. Let's aggregate to weekly totals using `floor_date()` to group days into weeks, then `summarize()` to add them up.

```{r}
weekly_wmata <- wmata |>
  mutate(week = floor_date(date, "week")) |>
  group_by(week) |>
  summarize(
    weekly_total = sum(total),
    days = n()
  ) |>
  filter(days == 7)  # Keep only complete weeks

head(weekly_wmata)
```

### Task 3: Make a line chart

Create a line chart of weekly ridership over time.

```{r}
weekly_wmata |>
  ggplot(aes(x = week, y = weekly_total)) +
  geom_line(color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Weekly WMATA Ridership (2024-2025)",
    x = "Week",
    y = "Total Riders"
  ) +
  theme_minimal()
```

**Answer this question:** Describe the trend you see. Is ridership going up, down, or staying flat? What are the big dips, and what might explain them? What would you want to know before writing a story about Metro weekly ridership trends?

Ridership is not staying consistent at all based on this chart, but it tends to hover around a certain amount beginning in the spring and slowly decline as it gets later into the year, with sharp dips at the beginning of each year. These dips could be explained by a number of things - bad weather or the holidays keeping people inside, for example - but I believe I remember seeing dips specifically around January 6th on both years in other datasets, which I chalked up to fear of anniversary protests over the insurrection from 2021. I would want to compare this data to ridership data from the same range in older years to find out if this is in line with annual trends or if 2024 saw unusually high or low ridership. Mainly, I want to know if COVID made ridership increase or decrease significantly once people began returning to public transportation after the pandemic. 
