---
title: "final_project"
output: html_document
---
###Setup
I just made a massive code block to load everything at once. Below I'm loading all the packages I want to use, as well as combining my spreadsheets so I can easily and cleanly compare all my data points. Since all the data exists on the same range, I just need to rename the columns so they don't become confusing when placed all together. 

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
library(corrplot)
library(stats)

visitor_volume <- read_csv("project/visitor_volume.csv") |> clean_names()
room_rates <- read_csv("project/average-room-rate.csv") |> clean_names()
casino_revenue <- read_csv("project/casino-revenue.csv") |> clean_names()

visitor_volume <- visitor_volume |> 
  mutate(month = mdy(month), yo_y_percent = parse_number(yo_y_percent)) |> 
  rename(visitors = monthly_value, visitor_yoy = yo_y_value, visitor_yoy_pct = yo_y_percent)

room_rates <- room_rates |> 
  mutate(month = date(mdy_hms(month)), yo_y_percent = parse_number(yo_y_percent)) |> 
  rename(roomrate = monthly_value, roomrate_yoy = yo_y_value, roomrate_yoy_pct = yo_y_percent)
  
casino_revenue <- casino_revenue |> 
  mutate(month = date(mdy_hms(month)), yo_y_percent = parse_number(yo_y_percent), yo_y_value = parse_number(yo_y_value), monthly_value = parse_number(monthly_value)) |> 
  rename(revenue = monthly_value, revenue_yoy = yo_y_value, revenue_yoy_pct = yo_y_percent)

visitor_plus_room <- visitor_volume |> 
  inner_join(room_rates, join_by(month))

combined <- visitor_plus_room |> 
  inner_join(casino_revenue, join_by(month))
```
###Analysis
I want to start by using a visualized correlation matrix to find out how gambling revenues and room rates interact with visitor rates. This will help me to find out which variables might have a correlation so I can dig into those further. 
```{r}
# Summary statistics
summary(combined)

# Create correlation matrix
selected_vars <- combined |> 
  select(visitors, roomrate, revenue)
correlation_matrix <- cor(selected_vars)
print(correlation_matrix)

# Visualize correlation matrix
corrplot(correlation_matrix, method = "circle")
```
Based on this matrix, there is a positive correlation between both variables in relation to visitor rates, although that seems more like a given for casino revenues than it does for room rates. I want to examine the latter further to see whether this correlation between rising room rates and rising visitor rates is significant and meaningful. Below, I'm calculating the pearson number for the correlation between the two and also measuring the linear regression model to see if the correlation is statistically significant. 

```{r}
#calculate the correlation
combined %>% summarize(cor(visitors, roomrate, method = "pearson"))
#calculate linear regression model
initial_model <- lm(visitors ~ roomrate, data = combined)
summary(initial_model)
```
The result of the pearson number shows that there is a positive relationship between 

something something paired t-test 
```{r}

pre_covid <- visitor_plus_room %>% filter(month < "2020-03-01")

pre_avg <- pre_covid %>% summarize(mean = mean(roomrate)) %>% pull(mean)

post_covid <- visitor_plus_room %>% filter(month>"2020-03-01")
post_avg <- post_covid %>% summarize(mean=mean(roomrate)) %>% pull(mean)

t.test(pre_avg, post_avg)
```
```{r}
visitor_plus_room |>
  ggplot(aes(x = `visitors`, y = `roomrate`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Visitors", y = "Room Rates",
       title = "Visitor rates vs. room pricing in Las Vegas")
```
This is really hard to read because of those outlier years of extremely low visitors. After taking a look at the data, it seems those data points are all months following the COVID-19 shutdown. The governor at the time, Steve Sisolak, was really strict on lockdown policy for a long time, so it continued to affect visitor rates well into 2021, which is what we're seeing here. But let's make a new scatter plot that better represents the rest of the data range. 
```{r}
library(dplyr)
library(ggplot2)

# Narrow the scatter plot 
filtered_data <- visitor_plus_room %>%
  filter(visitors >2e+06)

# Create the scatter plot with filtered data
  filtered_data %>% ggplot(aes(x = `visitors`, y = `roomrate`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Visitors", y = "Room Rates",
       title = "Visitor rates vs. room pricing in Las Vegas")
```
That's better! Although at face value, this is not nearly as clustered around the line of regression as one might hope. Just looking at this, you can certainly see the slight but significant positive relationship between visitor rates and room rates, meaning higher visitor numbers cause hotels to up their rates. 
```{r}
filtered_dates <- combined %>% filter(month > "2014-12-01")

#Make a timeplot of visitor rates over time
ggplot(filtered_dates, aes(x = month, y = visitors)) +
  geom_line(alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.1) +
  theme_minimal() +
  labs(title = "Monthly Visitors to Vegas Since 1980",
       x = "Date",
       y = "Visitors",
       caption = "Source: LVCVA")

#Make a timeplot of room rates over time 
ggplot(filtered_dates, aes(x = month, y = roomrate)) +
  geom_line(alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.1) +
  theme_minimal() +
  labs(title = "Vegas Hotel Room Rates Since 1980",
       x = "Date",
       y = "Average Room Rate",
       caption = "Source: LVCVA")

#Make a timeplot of gambling revenues over time 
ggplot(filtered_dates, aes(x = month, y = revenue)) +
  geom_line(alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.1) +
  theme_minimal() +
  labs(title = "Average Monthly Gambling Revenues Since 1980",
       x = "Month",
       y = "Average Gambling Revenue",
       caption = "Source: LVCVA")
```

