---
title: "JOUR405: Statistics for Journalists"
subtitle: "Midterm Exam"
name: "YOUR NAME HERE"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

## Load the tidyverse

```{r}
library(tidyverse)
```

---

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible, the compliance score, and the grade.

### Task 1: Read the data (2 points)

```{r}
moco_health <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```

### Task 2: Calculate the mean, standard deviation, minimum, and maximum of compliance scores. Save the results in a new dataframe called `compliance_stats`. (2 points for each)

```{r}
compliance_stats <- moco_health %>% summarize(
  (mean_score=mean(compliance_score)),
  (median_score=median(compliance_score)),
  (sd_score=sd(compliance_score))
)
```

### Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a descriptive title and label the x and y axes using `labs()`. (10 points)

```{r}
moco_health %>% 

ggplot(aes(x = compliance_score)) +
  geom_histogram(bins = 10, fill = "steelblue", color = "white") +
  geom_vline(aes(xintercept=mean(compliance_score)), color="red", linetype="dashed", linewidth = 1)+
  labs(
  title= "Distribution of Health Inspection Scores at Montgomery County Restaurants",
  x = "Score on health inspection",
  y = "Amount of restaurants"
  ) +
  theme_minimal()

```

### Task 4: Write 2–3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What is newsworthy, and what would your next reporting step be? (5 points)

The standard deviation score tells us that there are not many restaurants earning significantly higher or lower compliance scores than the mean, something which is also shown in the histogram. The mean rests right below where the majority of restaurants scored due to some of the lower scores bringing it down, but it is not far enough away to indicate any unusual deviation. Honestly, I don't find anything newsworthy about a lot of restaurants being really good at keeping up with health inspection scores. What I would want to do next is look into some of the restaurants that got bad scores and what the story is there, or look into the metrics for health inspection itself and see if they're being too lenient with the abundance of high scores. 

---

## Part 2: Daily Transit Ridership — Z-Scores and Sampling (30 points)

Your editor wants you to identify unusual days in WMATA's daily ridership data that might be worth investigating as story leads. The dataset contains daily bus and rail totals. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv`

### Task 5: Load the WMATA ridership data. (2 points)

```{r}
wmata_daily <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
```

### Task 6: Using `summarize()`, calculate the mean and standard deviation for both bus and rail ridership. Save the results in a dataframe called `wmata_stats`. (5 points)

```{r}
wmata_stats <- wmata_daily %>% summarize(
  mean_ridership=(mean(total)),
  sd_ridership=(sd(total))
)
```

### Task 7: Using `mutate()`, add two new columns to `wmata_daily`: `rail_zscore` (the z-score for rail ridership) and `bus_zscore` (the z-score for bus ridership). Save the result as `wmata_zscores`. (8 points)

```{r}
wmata_zscores <- wmata_daily %>% mutate(
    mean_rail = mean(rail),
    sd_rail = sd(rail),
    rail_zscore = (rail - mean_rail) / sd_rail,
    mean_bus = mean(bus),
    sd_bus=sd(bus),
    bus_zscore = (bus - mean_bus) / sd_bus
 )

```

### Task 8: Filter `wmata_zscores` to show only days where the absolute value of `rail_zscore` is greater than 2 (unusually high or low rail ridership). Arrange by the absolute value of the z-score, with the most extreme first. How many unusual days are there? (8 points)

```{r}
wmata_outliers <- wmata_zscores %>%
  filter(abs(rail_zscore) > 2) %>%
  arrange(desc(abs(rail_zscore)))
```

There are 6 total unusual days. 

### Task 9: Using `set.seed()`, take a random sample of 91 days from `wmata_daily` (roughly 25% of the data). Calculate the mean and standard deviation for bus and rail ridership in the sample. How do the sample statistics compare to the population statistics you calculated in Task 6? Write 1–2 sentences comparing the values and explaining what this suggests about random sampling. (4 points)

```{r}
set.seed(42)
wmata_sample <- wmata_daily %>% sample_n(91)

wmata_sample <- wmata_sample %>% mutate(
  mean_ridership=(mean(total)),
  sd_ridership=(sd(total)) 
)
```
The sample statistics have a lower mean ridership by about 20k and a higher standard deviation by about 15k. As we learned in class, smaller samples have a higher opportunity for variability, which likely accounts for the higher standard deviation in the smaller sample. The difference in means also goes to show that a higher polling population can change the mean to be more accurate. 

### Task 10: Write 2–3 sentences: What do the unusual days in your z-score analysis suggest as story leads? (3 points)

Most of the dates on the outlier list can be chalked up to Christmas, New Year's Day and Thanksgiving, all government holidays that would keep commuters from traveling. What's interesting are the two dates not attributed to holidays - both January 6th of different years. I would be interested in finding out whether that has to do with the anniversary of 1/6/21, and if local residents are afraid to commute that day for fear of a copycat attack or protest movement popping up. 

---

## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to find out which counties have the highest rates. The dataset contains car theft counts for Maryland counties in 2023 along with population figures. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv`

### Task 11: Load the car thefts data. (2 points)

```{r}
md_car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```

### Task 12: Using `mutate()`, add a column called `theft_rate` that calculates the car theft rate per 10,000 residents for each county. Save the result back to `md_car_thefts`. In 1–2 sentences, explain why a per-capita rate is more useful than raw counts for this comparison. (8 points)

```{r}
md_car_thefts <- md_car_thefts %>%  mutate(
  theft_rate = thefts_2023/population * 10000)
```

A per-capita rate is more useful when looking at thefts by county because counties in Maryland can have vastly different populations, and 100 thefts in a county with 100 people is a lot more significant than 100 thefts in a county with 100,000 people. By placing them in the context of the population, the rates are evened out. 

### Task 13: Calculate the median car theft rate and the total number of car thefts statewide. Show the counties that have rates above the median. (7 points)

```{r}
md_car_thefts <- md_car_thefts %>% mutate(
  median_rate = (median(theft_rate)),
)
car_theft_outliers <- md_car_thefts |>
  filter(abs(theft_rate) > 9.5) |>
  arrange(desc(abs(theft_rate)))

md_car_thefts<- md_car_thefts %>% mutate(
 total = nrow(thefts_2023)
)
```

### Task 14: Write 2–3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What is the lede of a story based on your findings? (8 points)

A significant amount of Maryland counties have a higher proportion of crimes than is the median for the state. I don't know anything about Maryland geography so I cannot speak to that, but PG County, St. Mary's County and Baltimore City in particular have much higher crime rates than the median. PG County and Baltimore City are obvious outliers in terms of crime and large urabn areas, but I think an interesting story could come out of why so much crime is happening proportionally in St. Mary's county as to put it next to these two city centers. 

## Part 4: Data Analysis Scenario (20 points)

A city parks department releases data showing that its 12 community swimming pools average 187 visitors per day during the summer season, with a standard deviation of 42 visitors. Your editor suspects that a handful of pools are dramatically underperforming and wants a story. You also learn that the department's figures are based on a random sample of 60 days per pool rather than a complete count.

Write 3–4 sentences for each of the following questions:

### Question A: Given that the mean is 187 and standard deviation is 42, which statistical measures would you prioritize using and why? (7 points)

I would prioritize looking at Z-scores for these pools to find out which pools are on the low end of the standard deviation, which is quite far from the mean. I would also want to know the median of the data, as the mean could be getting raised by dramatically popular pools. If the median is much lower than the mean, it could signify some underperforming pools. 

### Question B: How would you use z-scores to identify which pools are genuinely underperforming, and why is that more useful than simply ranking pools by raw attendance? (7 points)

I would look at the pools that have the most negative z-scores, which means they are further away from the mean on the low end. This is more helpful than just ranking the pools in a similar way to how looking at raw crime data is not helpful. Some of the pools might be smaller or in more rural areas; ranking them by their relation to the mean is much more helpful to see if they're underperforming on the whole than it would be to just look at their attendance without comparing it to the group mean. 

### Question C: The department collected its attendance figures using a random sample of 60 days per pool rather than tracking every day. What questions would you ask about the sample before relying on these figures for a story? What concerns would you raise in your story? Use statistical terminology where appropriate. (6 points)

60 days "per pool" is the part that sticks with me, although I also want to know the date range, as that obviously affects attendance. I want to know if these were consecutive days or 60 random days over the summer season, as well as if the same date ranges were used for each pool. If one pool was sampled in the early summer and another was sampled in late summer, their attendance statistics are hardly comparable. Before using these statistics, I would want to be clear on the methodology of this attendance polling and be able to include that in the story for full transparency about the statistics used. 

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your choices

Good luck!
