---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "Stella Garner"
---
```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

In the class activity, you tested a school board member's claim using data from Montgomery County, Pa. — and found a model that didn't hold up. Now you'll run the same analysis using data from Montgomery County, Md. Your job is to evaluate whether the Maryland data tells a different story, and what that means for how you'd report on the claim.

### Loading the Data

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```

### Task 1: Examine the Data First (2 points)

Before you plot anything, look at the data — you know why by now.

```{r}
summary(schools$`FARMS Pct`)
schools |> arrange(desc(`FARMS Pct`)) |> head(10)
```

Are there any data quality issues like the ones you found in the Pennsylvania data? How does the distribution of FARMS percentages compare between the two counties? Note anything that concerns you or gives you confidence in the data.

There are far less concerns for me just looking at this data on the surface, as there seems to be a normal incremental decrease in the FARMS eligibility at each school. However, there seems to be a similar random distribution of math scores just based on this sample, as I cannot follow any specific decreasing or increasing trend just looking at these schools. The highest number in proficiency belongs to the school with the lowest FARMS eligibility; the next highest number is the one with the highest. These are also vastly lower proficiency scores overall than we saw with the Pennsylvania data. 

### Task 2: Create a Scatterplot and Compare (5 points)

Replace the REPLACE_ME value with the correct variable name to create a scatterplot of FARMS percentage vs. math scores for Montgomery County, Md.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage, Montgomery County MD")
```

Describe what you see. Then compare this scatterplot specifically to the Pennsylvania one: what differences do you notice in the direction, strength, and spread of the relationship? The two counties share a name and are both suburban — what might explain why they look so different?

There does actually seem to be more of a shape to this data than what we saw with the Pennsylvania data, although it is still relatively scattered. It'll take some actual numbers to see whether there is noticeable improvement in statistical significance and correlation, but at face value, this chart seems much more likely to prove the board member's theory that schools with more students on FARMS see less proficiency in math testing. I did a little research on both counties, and MontCo (as the Pennsylvanians call it) has a AAA rating from S&P and about 300,000 less people than MoCo, which really makes me question where those 100 percent schools came from. Regardless, MoCo's lower testing scores overall and stronger correlation between variables could come from the fact that the county is much bigger and less well-off than MontCo. I am concerned, though, by the fact that it seems like MoCo eligibility for FARMS bottoms out at 5% rather than 0, and I'm curious if this is a minimum cutoff for the county. It's creating a line similar to, but not as dramatic as, the one from the MontCo data. 

### Task 3: Linear Regression Model (5 points)

Replace the REPLACE_ME value with the correct variable name to run a linear regression model predicting math scores from FARMS percentage.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```

Report the three key statistics: the p-value, the Adjusted R-squared, and the coefficient for FARMS Pct. Then interpret them together — not separately. What does the combination of these three numbers tell you about the relationship between poverty and math scores in Montgomery County, Md.? Is this model worth using? How does it compare to the Pennsylvania model?

The p-value here is almost incalculably small, meaning this data has really strong statistical significance. The adjusted R-squared value is about 67 percent, meaning over 2/3 of the data is impacted by the percentage of students on FARMS. The coefficient here is close to -60 percent, meaning there is a very strong negative correlation between FARMS eligibility and test scores, something we can already see on the scatter plot. This all adds up to prove the hypothesis that the school board member in Pennsylvania was attempting to make, albeit with a successful result this time; less students on FARMS means higher test scores in MoCo. I wouldn't say the model itself is actually better considering it's the exact same one, but that the data is telling a different story here based on this model - there's likely different factors that are larger indicators of test scores in Pennsylvania but don't mean so much in Maryland, and vice versa. 

### Task 4: Identifying Schools with the Biggest Residuals (6 points)

Replace the REPLACE_ME values with the correct variable names to calculate predicted values, residuals, and absolute residuals for each school. Then arrange the schools in descending order of absolute residuals.

```{r}
schools <- schools |>
  mutate(
    predicted = predict(model),
    residual = `% Met or Exceeded Math Expectations` - predicted,
    abs_residual = abs(residual)
  ) |>
  arrange(desc(abs_residual))

head(schools |> select(1, `FARMS Pct`, `% Met or Exceeded Math Expectations`, predicted, residual), 10)
```

Identify the school with the largest positive residual (doing much better than the model predicts) and the school with the largest negative residual (doing much worse). For each school, find its zip code and look it up on [CensusReporter.org](https://censusreporter.org).

For each of the two schools, answer:
- What does the demographic and economic profile of the zip code look like?
- Does anything in that profile help explain why the school is over- or under-performing the model?
- What would your first reporting question be to the principal of that school?

For Fox Chapel Elementary in Germantown, the median household income is over 100k and the poverty rate is 9.8 percent, which is lower than the rate in the U.S. overall. The strongest indicator of educational success for me in this zip code is that almost 9 in 10 adults in the area have a high school degree, and more than half hold a college degree. I would want to ask the principal why so many students in an area with such a high household income are eligible for FARMS, as it seems like this is why the model wasn't expecting the school to do as well as it did. If we were looking based on household income instead, I think these results would probably fall more in line. 

Germantown Elementary is not too drastically different - the median household income is lower at 86k (still not a small number), but the poverty rate is also lower by about a percentage point. However, only about 7 in 10 adults have high school degrees, with just 2 in 10 having college degrees. This is what I would predict accounts for the large disparity, although there are certainly other attributable factors that are not being recorded in this data. I would want to ask the principal if there are any large issues the school is dealing with that the data doesn't account for - a recent redistricting or curriculum overhaul, for example. The census numbers alone might account for the difference in the two schools, but not necessarily for how drastic of a difference Germantown is seeing.

### Task 5: Evaluate the Model and Write the Pitch (7 points)

You've now run this analysis on two counties. Return to the school board member's claim: that elementary schools with fewer students on free and reduced-price meals do better on standardized math tests.

Write a short story pitch of 4-6 sentences that addresses all of the following: Does the Maryland data support or contradict the claim? How does it compare to what you found in Pennsylvania? What would you ask the school board member if you called them for comment? And what's the most interesting specific finding from your residuals analysis: which school would you want to visit first, and why?

The Maryland data supports the board member's claim that schools with higher FARMS eligibility tend to have lower test scores. This is a stark contrast to the data in Pennsylvania, which found no such correlation and even trended slightly in the opposite direction. If calling the school board member for comment, I would want to ask what data or resource brought them to make this claim - maybe they got their counties mixed up. The information on Fox Chapel elementary is interesting to me, not in that they have surprisingly high test scores but in that so many of their students are eligible for FARMS considering the census data shows it to be an incredibly affluent community. My first order of business would be to visit the school and find out what exactly is going on with FARMS and if there's something about the student population the census data or the data in the model isn't reflective of. 
