---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2025.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

The first thing which sticks out to me, aside from the fact that the "pay_basis" column seems pretty moot to have here, are the few people who are listed on the payroll with absolutely no salary. Not only will this drag down the mean and median in analysis, but it also begs the question of why those staffers are listed on the payroll at all - especially considering some of the bizarre titles they hold, like "crypto czar" and "senior advisor to the White House Faith Office."

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries %>%  summarize(mean_salary=mean(SALARY))
wh_salaries %>%  summarize(median_salary=median(SALARY))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries %>% summarize(min_salary=min(SALARY))
wh_salaries %>% summarize(max_salary=max(SALARY))
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean and median salary are relatively close in number, meaning that the salaries are somewhat evenly distributed and there are not any huge outliers with the exception of the unsalaried listed workers. Again, it is puzzling as to why these individuals would be included, something which could be a story of its own. The only two salaries on the list over $200k are for an advisor and a senior counsel, which I also find interesting because they are paid more than seemingly high-ranking positions like Chief of Staff. 

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

I would be interested in learning more about the two individuals who are paid more than $200k and what they do to make more than Susie Wiles. I would likely do some research into who each individual is and what their job theoretically entails, as well as what sentiment around them in Washington is. I would also want to look into the historical pay range for these positions to see if this is a normal salary range or if this administration is paying them more richly than the historical precedent. 
