---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv") %>% clean_names()
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities %>% mutate(violent_rate_2019=violent2019/pop2019 * 500, property_rate_2019=property2019/pop2019 * 500, violent_rate_2020=violent2020/pop2020 * 500, property_rate_2020=property2020/pop2020 * 500)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

There is not a very consistent increase or decrease in overall crime rates across cities, though it does seem that a general decrease in crime is more common overall. Certain cities did see particular spikes or have oddly high statistics - the property crime rates in Ocean City, for example, which make sense when you consider the type of people living in and visiting the area. In the city of Somerset, property crime rates more than doubled despite a decrease in population. North Chevy Chase, on the other hand, saw property rates more than cut in half from 2019 to 2020.   

### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates %>% mutate(property_change=(property_rate_2020-property_rate_2019)/property_rate_2019 *100)
```


```{r}
md_cities_change <- md_cities_change %>% mutate(violent_change=(violent_rate_2020-violent_rate_2019)/violent_rate_2019 *100)
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

Most of the drastic percentage changes are owed to a small number of crimes or a small population changing in a way that is substantial relative to the total number, but not overall. One that did stick out to me is the 320 percent increase in property crimes in Crisfield, from 3 in 2019 to 46 in 2020. On the other side of the spectrum are large cities which saw a sizeable change in their crime rates in number but not necessarily percentage, such as the drop from 575 property crimes to 378 in 2019.  

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

2020 was a year of decreases for many things - travel, gatherings and so on - but in Crisfield, one thing saw a stark increase: property crimes. 

From 2019 to 2020, the Somerset County town saw property crime rates jump over 300 percent, with the rate of property crimes per 500 people rising from about 2 to nearly 10 in 2020. 

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

### Task 10

```{r}
md_cities_percentiles <- md_cities_rates %>% mutate(percentile = ntile(violent_rate_2020, 100))
```
There are 8 cities above the 90th percentile of violent crime rates in 2020. 
```{r}
md_cities_percentiles %>% summarize(IQR=IQR(violent_rate_2020))
```
Based on the small value of this IQR, I get the sense that most crime rates in Maryland are pretty middling. Although this is a good thing, I think it would be a difficult statistic to turn into an interesting story unless you were the press secretary for Maryland State Police. 
