---
title: "HW8b_PollingSimulation"
output: html_document
---

In class, you simulated polling estimates for the share of Americans who are religiously unaffiliated — a group that makes up about 36% of the population. In this assignment, you'll run the same simulation for a different group: White Evangelicals, who make up roughly 10% of the population.

That difference matters more than it might seem. When a proportion is close to 50%, samples tend to be pretty well-behaved. When it's much smaller — or much larger — samples can behave differently, especially at small sizes. Your job is to run the simulation, compare what you find to the in-class results, and think carefully about what it means for reporting on religion and politics.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the CES data

Load the CES 2024 religion dataset from the same URL as the in-class activity and save it to a dataframe called `ces`:

`https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv`

```{r}
ces <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv")
```

### Task 3: Calculate the population proportion

Create a new column called `evangelical` that is `TRUE` if a respondent's `trad` value is `"White Evangelical"` and `FALSE` otherwise. Then calculate the proportion of the full dataset that is White Evangelical and save it as `population_proportion`.

```{r}
 ces <- ces %>% 
  mutate(evangelical = trad %in% c("White Evangelical"))
population_proportion <- mean(ces$evangelical)

```

Write down this value. How does it compare to the unaffiliated proportion (about 0.36) from the in-class activity?

The value rounded to the hundreds is 0.10. This is a far lower value than the unaffiliated proportion, meaning there are far fewer evangelicals in proportion to the full population than there are unaffiliated people. 

### Task 4: Take a single small sample

Use `set.seed(42)` and sample 50 rows from `ces`. Save it as `sample_50`. Calculate the proportion of White Evangelical respondents in your sample.

```{r}
set.seed(42)
sample_50 <- ces %>% sample_n(50)
proportion_50 <- mean(sample_50$evangelical)
```

How close is your sample estimate to the population proportion? How does the accuracy compare to what you saw with the unaffiliated group in class?

The sample estimate is a little bit higher than the population proportion, at 0.16. This is a little bit different from the unaffiliated group in the sample from class, which actually saw a somewhat higher proportion in the sample than it did in the population. 

### Task 5: Take a single larger sample

Use `set.seed(1966)` and sample 500 rows. Save it as `sample_500`. Calculate the proportion of White Evangelical respondents in this sample.

```{r}
set.seed(1966)
sample_500 <- ces %>% sample_n(500)
proportion_500 <- mean(sample_500$evangelical)
```

Did the larger sample improve the estimate? How does the improvement compare to what you saw in class?

The larger sample did get the estimate closer, though it actually brought it below the proportion in the full population. This actually made me realize that I made a mistake in my class assignment by not changing the set seed, which is why I ended up with the same proportions in that case. Whoops. The estimate probably would've gotten closer in that case too, if I had to guess. 

### Task 6: Simulate many polls

The code below runs the same simulation as the in-class activity, but using your `evangelical` column. Run it as-is.

```{r}
set.seed(42)

simulate_polls <- function(data, sample_size, n_simulations = 1000) {
  proportions <- replicate(n_simulations, {
    s <- data |> sample_n(sample_size)
    mean(s$evangelical)
  })
  tibble(proportion = proportions, sample_size = sample_size)
}

simulations_50 <- simulate_polls(ces, 50)
simulations_500 <- simulate_polls(ces, 500)
```

### Task 7: Visualize the simulation results

Create two histograms side by side using the code structure from class. Add a vertical line at `population_proportion`. Make sure to update the title and subtitle to reflect that this simulation is about White Evangelicals.

```{r}
simulations_all <- bind_rows(simulations_50, simulations_500) |> 
  mutate(sample_size = paste("n =", sample_size))

ggplot(simulations_all, aes(x = proportion)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept=population_proportion, color="red", linetype="dashed", linewidth = 1)+
  facet_wrap(~sample_size) +
  labs(
    title = "Distribution of Sample Estimates: Share of Evangelicals",
    subtitle = "Red line = true population proportion. Each histogram shows 1,000 simulated polls.",
    x = "Sample Proportion",
    y = "Count",
    caption = "Source: 2024 Cooperative Election Study"
  ) +
  theme_minimal()
```

### Task 8: Reflect on what you see

**Describe the shape of the n=50 histogram.** Is it as symmetric as the unaffiliated histogram from class? Why do you think that is, given that White Evangelicals make up only about 10% of the population?

It has relatively the same shape and symmetry as the unaffiliated n=50 histogram, albeit a little more spaced out. I could be very wrong here, but don't think it actually has anything to do with the proportions of the percentages - because we are still testing 1000 simulations, there is bound to be this amount of range in the proportions we receive back from these samples. 

**Compare the n=50 and n=500 histograms for this group.** How much does the distribution change with a larger sample? Is the improvement bigger or smaller than what you saw in class with the unaffiliated group?

The distribution is far more clustered around the mean in the n=500 histogram, showing a flattening of outliers as sample sizes get larger. I would say the improvement is relatively the same compared to the imrprovement from class, although there is a higher number of sample proportions higher than the mean instead of lower. 

**A news organization wants to report on White Evangelical voters using a poll of 200 people.** Based on your simulations, roughly how many White Evangelical respondents would that poll be likely to include? What does that suggest about the reliability of estimates for smaller subgroups?

Based on our population proportion of 0.10, a poll of 200 would likely include about 20 evangelical voters. This suggests that a larger sample size would always be better for estimating small subgroups through polling, as the larger your sample size is, the more people from that subgroup you can include. Small sample sizes will not be reliable for polling small groups because you will not be able to talk to a representative amount of the demographic. 

**Imagine you are covering the 2028 presidential election and a campaign shares a poll showing White Evangelical support at 78%.** The poll sampled 400 people. Based on everything you've seen in this assignment and in class, what would you want to know before reporting that number?

I would want to know how many of those people identified as Evangelical and what the other groups included in the survey were. Based on our proportion, this means that likely around 40 of the people polled consider themselves Evangelicals, IF this is a survey of the general population. If that is the case, I would not want to report the number because that is not a large enough group to reliably say they represent the whole of the Evangelical population. However, if the study was done among 400 Evangelicals or among 400 people in an Evangelical stronghold, I would be more likely to report that number knowing the population proportion is likely higher in those groups and therefore more indicative of the demographic's opinions. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
