---
title: "final_project"
output: html_document
---
###Setup
I made a massive code block to load everything at once. Below I'm loading all the packages I want to use, as well as combining my spreadsheets so I can easily and cleanly compare all my data points. Since all the data exists on the same range, I just need to rename the specific data columns so they don't become confusing when placed all together. 

```{r}
#Load it up
library(tidyverse)
library(lubridate)
library(janitor)
library(corrplot)
library(stats)

#Create cleaner data tables 
visitor_volume <- read_csv("project/visitor_volume.csv") |> clean_names()
room_rates <- read_csv("project/average-room-rate.csv") |> clean_names()
casino_revenue <- read_csv("project/casino-revenue.csv") |> clean_names()

#Create different combinations of data 
visitor_volume <- visitor_volume |> 
  mutate(month = mdy(month), yo_y_percent = parse_number(yo_y_percent)) |> 
  rename(visitors = monthly_value, visitor_yoy = yo_y_value, visitor_yoy_pct = yo_y_percent)

room_rates <- room_rates |> 
  mutate(month = date(mdy_hms(month)), yo_y_percent = parse_number(yo_y_percent)) |> 
  rename(roomrate = monthly_value, roomrate_yoy = yo_y_value, roomrate_yoy_pct = yo_y_percent)
  
casino_revenue <- casino_revenue |> 
  mutate(month = date(mdy_hms(month)), yo_y_percent = parse_number(yo_y_percent), yo_y_value = parse_number(yo_y_value), monthly_value = parse_number(monthly_value)) |> 
  rename(revenue = monthly_value, revenue_yoy = yo_y_value, revenue_yoy_pct = yo_y_percent)

visitor_plus_room <- visitor_volume |> 
  inner_join(room_rates, join_by(month))

combined <- visitor_plus_room |> 
  inner_join(casino_revenue, join_by(month))
```
###Analysis
I want to start by using a visualized correlation matrix to find out how gambling revenues and room rates interact with visitor rates. This will help me to find out which variables might have a correlation so I can dig into those further. 

```{r}
# Summary statistics
summary(combined)

# Create correlation matrix
selected_vars <- combined |> 
  select(visitors, roomrate, revenue)
correlation_matrix <- cor(selected_vars)
print(correlation_matrix)

# Visualize correlation matrix
corrplot(correlation_matrix, method = "circle")
```
Based on this matrix, there is a positive correlation between both variables in relation to visitor rates, although that seems more like a given for casino revenues than it does for room rates. I want to examine the latter further to see whether this correlation between rising room rates and rising visitor rates is significant and meaningful. 

Below, I'm calculating the pearson number for the correlation between the two and also measuring the linear regression model to see if the correlation we've just established is statistically significant. 

```{r}
#calculate the correlation
combined %>% summarize(cor(visitors, roomrate, method = "pearson"))
#calculate linear regression model
initial_model <- lm(visitors ~ roomrate, data = combined)
summary(initial_model)
```
The result of the pearson number shows that there is a moderately positive relationship between visitor rates increasing and room rates increasing. I will want to do some more testing later to determine which factor is influencing the other. The linear regression model indicates that the model is extremely statistically significant, meaning these results could not be due to chance; however, the multiple and adjusted R-squared values indicate that a model based on this correlation would not be particularly strong or sound. 

I also noticed that some of the biggest shifts in visitor rates and pricing took place during the COVID pandemic; I want to split my data into pre- and post-COVID blocks to see if that affects what my regression charts look like. I'll also use a t-test to see if there's a drastic different between the means of those two blocks of data. 
```{r}
#Create data blocks for pre- and post-COVID
pre_covid <- visitor_plus_room %>% filter(month < "2020-03-01")
post_covid <- visitor_plus_room %>% filter(month>"2020-03-01")

#Find the averages 
pre_avg <- pre_covid %>% summarize(mean = mean(roomrate)) %>% pull(mean)
post_avg <- post_covid %>% summarize(mean=mean(roomrate)) %>% pull(mean)

#Create the t-test
t.test(pre_covid$roomrate, post_covid$roomrate, alternative = "two.sided")
```
While this t-test does clearly show a drastic difference between the two room rate means, I also want to find out if the mean of visitor rates rose accordingly to meet the rise in room rates. 

```{r}
t.test(pre_covid$visitors, post_covid$visitors, alternative = "two.sided")
```
This is really interesting - where room rates rose post-COVID, visitor rates shrunk. I'll keep this in mind for my story pitch later. Now I want to dig more into the correlation between visitor rates and room rates by visualizing the data - I'll start with just a scatterplot of all the data available. 
```{r}
visitor_plus_room |>
  ggplot(aes(x = `visitors`, y = `roomrate`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Visitors", y = "Room Rates",
       title = "Visitor rates vs. room pricing in Las Vegas")
```
This is really hard to read because of those outlier years of extremely low visitors. After taking a look at the data, it seems those data points are all months following the COVID-19 shutdown. The governor at the time, Steve Sisolak, was really strict on lockdown policy for a long time, so it continued to affect visitor rates well into 2021, which is what we're seeing here. But let's make a new scatter plot that better represents the rest of the data range. 

```{r}
library(dplyr)
library(ggplot2)

# Narrow the scatter plot 
filtered_data <- visitor_plus_room %>%
  filter(visitors >2e+06)

# Create the scatter plot with filtered data
  filtered_data %>% ggplot(aes(x = `visitors`, y = `roomrate`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Visitors", y = "Room Rates",
       title = "Visitor rates vs. room pricing in Las Vegas")
```
That's better. Although at face value, this is not nearly as clustered around the line of regression as one might hope. Just looking at this, you can certainly see the slight but significant positive relationship between visitor rates and room rates, meaning higher visitor numbers cause hotels to up their rates. 

Considering that we've already split up our data into pre- and post-covid blocks and found some interesting results based on that, let's also take a look at what scatterplots specifically based on those time periods look like. I want to find out if the data will be more strongly correlated during a certain time period, and whether COVID may have affected that correlation. 

```{r}

# Create the pre-COVID scatter plot
  pre_covid %>% ggplot(aes(x = `visitors`, y = `roomrate`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Visitors", y = "Room Rates",
       title = "Visitor rates vs. room pricing in Las Vegas, Pre-COVID")

# Create the post-COVID scatter plot 
  post_covid %>% ggplot(aes(x = `visitors`, y = `roomrate`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Visitors", y = "Room Rates",
       title = "Visitor rates vs. room pricing in Las Vegas, Post-COVID")

```
Obviously these are difficult data points to compare considering the number of them is so vastly different between the two time periods. However, it does appear that the correlation remains similar after the COVID pandemic, even when accounting for months from 2020 to 2021, when almost no visitors were recorded; hotel room rates were accordingly very low. Let's also check the pearson number for both to see if either one has a stronger correlation. 

```{r}
#Calculate pre-COVID pearson number 
pre_covid %>% summarize(cor(visitors, roomrate, method = "pearson"))

#Calculate post-COVID pearson number 
post_covid %>% summarize(cor(visitors, roomrate, method = "pearson"))
```
My suspicions based on the scatter plots are mainly confirmed; the correlations are close to the same in both cases. In fact, the correlation between rising rates and rising visitors is higher post-COVID, although this could be due to the very low visitors/room rates which occurred during the pandemic months. But I also think it's possible that hotels felt justified in raising prices higher post-COVID than pre-COVID because there was a lack of recent precedent for the prices; this is something I'll address in my story pitch. 

All this data lends to finding out whether the hypothesis I had going into this is true, which is that high room rates were one of the potential contributors to the decline in visitors that Vegas experienced in 2025. First, I want to look at how each dataset has changed over time, so I'm putting every dataset on a different timeline. I'm going to start by filtering the data just to show trend lines starting in 2015 so we can look at the last decade specifically and see if anything stands out. 
```{r}
filtered_dates <- combined %>% filter(month > "2014-12-01")

#Make a timeplot of visitor rates over time
ggplot(filtered_dates, aes(x = month, y = visitors)) +
  geom_line(alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.1) +
  theme_minimal() +
  labs(title = "Monthly Visitors to Vegas Since 2015",
       x = "Date",
       y = "Visitors",
       caption = "Source: LVCVA")

#Make a timeplot of room rates over time 
ggplot(filtered_dates, aes(x = month, y = roomrate)) +
  geom_line(alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.1) +
  theme_minimal() +
  labs(title = "Vegas Hotel Room Rates Since 2015",
       x = "Date",
       y = "Average Room Rate",
       caption = "Source: LVCVA")

#Make a timeplot of gambling revenues over time 
ggplot(filtered_dates, aes(x = month, y = revenue)) +
  geom_line(alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.1) +
  theme_minimal() +
  labs(title = "Average Monthly Gambling Revenues Since 2015",
       x = "Month",
       y = "Average Gambling Revenue",
       caption = "Source: LVCVA")
```
An interesting pattern here is how all of the data seems to have seasonal trendlines that mainly stayed the same up until COVID. It might be interesting just to go into the ways that COVID genearlly affected all of these variables in my story pitch. The dropoff in monthly visitors around early 2025 is fairly visible, but I want to emphasize it so I can really compare it to gambling revenues and hotel room rates. Like I did with the scatterplots, I'll narrow down the data into the time period I'm interested on so I can really look at those trend lines in detail. Since it's such a small amount of data and the points are exact, the geom_smooth function will not be necessary here. 
```{r}
year_dates <- combined %>% filter(month > "2024-12-01")

#Make a 2025 timeplot of visitor rates
ggplot(year_dates, aes(x = month, y = visitors)) +
  geom_line(alpha = 0.5) +
  theme_minimal() +
  labs(title = "Monthly Visitors to Vegas in 2025",
       x = "Date",
       y = "Visitors",
       caption = "Source: LVCVA")

#Make a 2025 timeplot of room rates 
ggplot(year_dates, aes(x = month, y = roomrate)) +
  geom_line(alpha = 0.5) +
  theme_minimal() +
  labs(title = "Vegas Hotel Room Rates in 2025",
       x = "Date",
       y = "Average Room Rate",
       caption = "Source: LVCVA")

#Make a 2025 timeplot of gambling revenues
ggplot(year_dates, aes(x = month, y = revenue)) +
  geom_line(alpha = 0.5) +
  theme_minimal() +
  labs(title = "Average Monthly Gambling Revenues in 2025",
       x = "Month",
       y = "Average Gambling Revenue",
       caption = "Source: LVCVA")
```
This is interesting. The visitor rates for this time period, when zoomed in on, are a little all over the place, with a stark dropoff shortly after January that could be attributed to immediate effects of President Trump assuming office and international tourists choosing not to enter the United States. All of the charts do show a drastic crater during the summer months, although this is less pronounced in visitor rates themselves and more pronounced in room rates and gambling revenues. Although summer usually tends to be a little slower based on our seasonal data from the previous time plots, in none of them do rates drop so drastically like they do here. 

###Story Pitch 
My main takeaway from these findings is that, while there does appear to be a statistically significant correlation between a rise in visitor rates and a rise in room rates, it is neither newsworthy nor interesting enough to be reported on further. But I do think there are a couple more areas in this data where interesting stories might lie. One lies with how room rates have existed on an entirely new, increased precedent since visitor rates bounced back after COVID, and they don't seem likely to drop back to pre-COVID rates anytime soon.
Based on our t-tests, I find it possible this new price precedent is part of the reason why post-COVID visitor means dipped a little. I also think the large crater in specifically the data from the summer months of 2025 would be interesting to explore. If visitor rates actually increased slightly in the preceding months, why would they have dropped so much more drastically than what is typical for the summer season? 

To explore both of these ideas further, I would start by looking into the history of specific hotel's room rates both pre- and post-COVID and talk to hotels who raised their prices the most significantly to find out why that might have been the case. I also would want to talk the regular revisitors that make up a large part of the Vegas tourism base - have these rising room rates discouraged them from returning or affected the nature of their trips at all, and could that have contributed to the dip seen in the summer of 2025? Lastly, I would want to speak to other business owners who might have been affected by that summer dip. Have they struggled to get by more frequently in recent years during those down periods, or have rising prices contributed to higher revenues? And what was the experience of that specific summer like? 
